Improving the Accuracy of a Sequential Mining Model for Hurricane Trajectory Prediction

### 2017-12-06 paper/manuscript.tex
little tex fix
Complete Problem Description and start Soluiton section

\usepackage{graphicx}
\usepackage{textcomp}
\graphicspath{ {images/} }

@@

\begin{figure}[h]
\caption{Sample of mapped Hurricane-Coordinates database}
\centering
\includegraphics[scale=1.0]{hurricane-table-sample}
\end{figure}

The above table shows an example of how the trajectory database is structured after the mapping preprocessing step. Only two dimensions are needed in the database, being the hurricane ID and the sequence of latitude-longtitude coordinates. All hurricanes in HURDAT2 have an ID corresponding to the location (the Atlantic), the index ordered by appearance in the hurricane year, and the year itself. For example, hurricane Katrina of 2005 corresponds to AL122005. The preprocessing step proposed is executed exactly once, allowing the hurricane database to be mined without the initial grouping and mapping stage.

@@

Section (2) of AprioriAll initiates the generation of frequent sequences. In the case of hurricane trajectory prediction, these sequence correspond to frequent recorded hurricane coordinate sequences. Frequent hurricane coordinate sequences are those which occur at least a minimum number of times, as defined by the \textit{minimum support}, or minsup. In general terms, the sequence generation step of AprioriAll develops frequent ordered patterns of increasing length, beginning with frequent single-item sequences (corresponding to frequent single-point sequences). This step begins by scanning the $\langle HurricaneID, CoordinateSequence\rangle$ database for frequent single-point sequences. The frequent single-point trajectories are then converted to candidate two-point trajectories as the result of an inner join. The candidate two-point trajectories are scanned against the database to determine which candidates satisfy the minimum support. Another inner join generates the candidate 3-length trajectories. Once the sequence generation level reaches three, each candidate sequence must be pruned if any of their candidate subsequences are infrequent. Continuous execution of AprioriAll until completion results in a collection of frequent hurricane trajectory patterns, as defined by the minumum support.

@@

Consider figures (2) and (3) shown above, generated using a sample of the recrded hurricane trajectories in HURDAT2 with the visualization tool Matplotlib \cite{matplotlib}. As shown in the figures, region discretization organizes continuous hurricane coordinate sequences into discrete region sequences of approximately uniform size and square shape. A discretization block size of 5 is selected in HTPDM and is replicated in WARD-HTP. A block size of 5 corresponds approximately to a real-world block dimension of 500km by 500km. The block dimension size is approximate because the mapping between coordinate degrees and kilometers is imprecise, and is dependent on location in the case of longtitudal points \cite{lat-long-distance}. The approximate size of 500km is similar to the average hurricane total diameter \cite{hurricane-distances}, permitting the coarse-grain expression of hurricane trajectories without undue loss in precision.

@@

% this may need some editing

@@

\subsection{Haversine Formula}

@@

% Early callouts:
% Motivation could be more clear
% We need to make sure we are explaining rule-miss incidents and our approach to them
% We also need to make sure we explain what exactly the correctness ratio is and how it changes based on accounting for rule-miss
% future work: non-binary prediction correctness?
% More non-text assets?
% We have done a good job describing the problem space but we need to explain what we problems __are__
% ie: why is there a solution

>>>>

\usepackage{graphicx}
%\usepackage{textcomp}
\graphicspath{ {images/} }

@@

\begin{figure}[ht]
\caption{Sample of mapped Hurricane-Coordinates database}
\centering
\includegraphics[scale=1.0]{hurricane-table-sample}
\end{figure}

Figure (1) shows an example of how the trajectory database is structured after the mapping preprocessing step. Only two dimensions are needed in the database, being the hurricane ID and the sequence of latitude-longtitude coordinates. All hurricanes in HURDAT2 have an ID corresponding to the location (the Atlantic), the index ordered by appearance in the hurricane year, and the year itself. For example, hurricane Katrina of 2005 corresponds to AL122005. The preprocessing step proposed is executed exactly once, allowing the hurricane database to be mined without the initial grouping and mapping stage.

@@

Section (2) of AprioriAll initiates the generation of frequent sequences. In the case of hurricane trajectory prediction, these sequences correspond to frequent recorded hurricane coordinate sequences. Frequent hurricane coordinate sequences are those which occur at least a minimum number of times, as defined by the \textit{minimum support}, or minsup. In general terms, the sequence generation step of AprioriAll develops frequent ordered patterns of increasing length, beginning with frequent single-item sequences (corresponding to frequent single-point sequences). This step begins by scanning the $\langle HurricaneID, CoordinateSequence\rangle$ database for frequent single-point sequences. The frequent single-point trajectories are then converted to candidate two-point trajectories as the result of an inner join. The candidate two-point trajectories are scanned against the database to determine which candidates satisfy the minimum support. Another inner join generates the candidate 3-length trajectories. Once the sequence generation level reaches three, each candidate sequence must be pruned if any of their candidate subsequences are infrequent. Continuous execution of AprioriAll until completion results in a collection of frequent hurricane trajectory patterns, as defined by the minumum support.

@@

Consider figure (2), generated using a sample of the recrded hurricane trajectories in HURDAT2 with the visualization tool Matplotlib \cite{matplotlib}. As shown in the pair of images, region discretization organizes continuous hurricane coordinate sequences into discrete region sequences of approximately uniform size and square shape. A discretization block size of 5 is selected in HTPDM and is replicated in WARD-HTP. A block size of 5 corresponds approximately to a real-world block dimension of 500km by 500km. The block dimension size is approximate because the mapping between coordinate degrees and kilometers is imprecise, and is dependent on location in the case of longtitudal points \cite{lat-long-distance}. The approximate size of 500km is similar to the average hurricane total diameter \cite{hurricane-distances}, permitting the coarse-grain expression of hurricane trajectories without undue loss in precision.

@@

The hurricane trajecory prediction model based on data mining as originally proposed in \cite{major-hurricane-model} implements all aspects of the problem domain described. Modified AprioriAll, downscaled region discretization and pattern matching with respect to association rule antecedents and consequents are respectively employed by HTPDM in order to develop an effective system for predicting future hurricane trajectories. Based on Atlantic hurricane trajectory data from 1900 to 2008, the model generates a best-case correctness ratio of 65\% and a worst-case ratio of 57.5\%. The training data used in the original model consists of trajectories from 1900 to 2000, with testing data supplied by trajectories from 2001 to 2008. At a high level, the model is developed and tested by following the steps below, using all implementation strategies discussed.

% Model formal explanation could be better. Maybe do pseudocode

\begin{enumerate}
\item Generate interesting trajectory rules from AprioriAll
\item Determine matching rule for incoming trajectory 
\item Check if the matched rule's consequent points are sufficiently close
\end{enumerate}

The differentiation between the model's best and worst-case hit rates is dependent on the ability of an incoming test trajectory to be matched with a trajectory association rule. As described in the pattern matching stage, the initial subtrajectory of an incoming trajectory is compared against all interesting rules to determine a best-fit match. If no such match is found, HTPDM attempts to extrapolate a future point of the initial trajectory, using the extrapolated point as a new option for determing a rule antecedent match. When accounting for all tested trajectories, including those which were matched to a rule as a result of trajectory extrapolation, the correctness ratio was 57.5\%. Therefore, the best case correctness ratio of HTPDM is the ratio which ignores all test trajectories which did not match with any rules, resulting in a final value of 65\%.

Despite establishing a convincing initial start point, the best-case correctness ratio of 65\% may be too low for HTPDM to be considered realiable in the real-world. This is problematic for those interested in hurricane trajectory prediction as a model with only 65\% accuracy may not be able to deliver crucial insight into hurricane trajectory prediction, particularly with respect to landfall location prediction. Both correctness ratios given must be improved in order for data mining to be considered a reliable approach to predicting the trajectories of future hurricanes.

@@

In order to improve the correctness ratio of datamining-based hurricane trajectory prediction, this work identifies and explores three opportunities for worst-case and best-case ratio improvement. The opportunities for improvement are the improvement of realism, reduction of historical impact, and favorability towards recent hurricane trends. Therefore, the opportunities for improvement are realized by three unique dimensions. These dimensions are the use of a distance formula designed for spherical coordinate systems, the Haversine Distance Formula, the use of weighted training data when executing the modified variant of AprioriAll, and the modification of the training data to only use historical trajectories from 1950 to 2000. The implementation of these three dimensions results in a new hurricane trajectory prediction model dubbed WARD-HTP (Weighted-Asset, Realistic-Distance Hurricane Trajectory Predictor). The combined contributions of WARD-HTP are expected to result in improvements to the best-case and worst-case correctness ratios.

\subsection{Haversine Distance Formula}

@@

% Early callouts:
% Motivation could be more clear
% We need to make sure we are explaining rule-miss incidents and our approach to them
% We also need to make sure we explain what exactly the correctness ratio is and how it changes based on accounting for rule-miss
% Also need to explicitly call out what our minconf was and why
% future work: non-binary prediction correctness?
% More non-text assets?
% We have done a good job describing the problem space but we need to explain what we problems __are__
% ie: why is there a solution
% we will need to re-read the paper and keep an eye out for sections that suffer from scope creep. Problem description first comes to mind.
